---
title: "Getting Started with ecsr10"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
  pdf_document:
    toc: true
    toc_depth: 2
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Getting Started with ecsr10}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

The `ecsr10` package provides an R implementation of the Elixhauser Comorbidity Software Refined (CMR) workflow for analyzing ICD-10-CM diagnosis codes and calculating comorbidity indices. This package is designed for healthcare researchers and data analysts who need to identify comorbidities and calculate risk scores from patient diagnosis data.

It implements AHRQ/HCUP releases **v2022.1 through v2026.1**, selectable per call. The default, `cmr_version()`, is v2026.1, covering ICD-10-CM diagnosis codes from October 2015 through September 2026 (ICD-10-CM versions 33–43). Record the release alongside any published results: the same patient can flag differently across releases, and reproducing an existing cohort means scoring it under the release it was built with.

## Installation

Install the development version from GitHub:

```{r eval=FALSE}
# install.packages("devtools")
devtools::install_github("dpei/ecsr10")
```

```{r setup}
library(ecsr10)
library(dplyr)

cmr_releases()   # every release this package can score with
cmr_version()    # the one used when you don't say
```

## Basic Workflow

The ecsr10 package provides a simple workflow for most users:

1. **Prepare your patient diagnosis data** (wide format with diagnosis and POA columns)
2. **Apply comorbidity analysis** using the built-in lookup data
3. **Calculate risk indices** for readmission and mortality

The package automatically uses built-in lookup data, making the process straightforward.

## Step 1: Data Preparation

Your patient diagnosis data should be in wide format with:
- Diagnosis code columns (DX1, DX2, ..., DXn)
- Present on Admission (POA) columns (POA1, POA2, ..., POAn)  
- Year and quarter for ICD version determination
- Patient/encounter identifiers

```{r}
# Example patient diagnosis data structure
example_patient_data <- data.frame(
  encounter_id = 1:3,
  test_type = "example",
  dx1 = c("E119", "I10", "Z5111"),
  dx2 = c("I10", "E119", "N183"),
  dx3 = c(NA, "N183", NA),
  poa1 = c("Y", "Y", "Y"),
  poa2 = c("N", "Y", "W"),
  poa3 = c(NA, "Y", NA),
  year = 2025,
  quarter = 1,
  stringsAsFactors = FALSE
)

print(example_patient_data)
```

## Quick Start Example

```{r}
# Simple workflow - uses built-in lookup data automatically
result <- comorbidity(example_patient_data, 
                      dx_cols = c("dx1", "dx2", "dx3"),
                      poa_cols = c("poa1", "poa2", "poa3"))

# Calculate risk indices
result_with_indices <- cmr_index(result)

# View summary
head(result_with_indices[,c("encounter_id", "CMR_Index_Readmission", "CMR_Index_Mortality")])
```

That's it! The package handles all the complex lookup building internally.

### Using multiple cores

`comorbidity()` takes an `ncores` argument (default `1`) that splits the
encounters into that many contiguous blocks and runs the whole pipeline on each
block in a forked worker:

```{r eval=FALSE}
result <- comorbidity(example_patient_data,
                      dx_cols = c("dx1", "dx2", "dx3"),
                      poa_cols = c("poa1", "poa2", "poa3"),
                      ncores = 4)
```

The result is identical for every `ncores` value. Every stage of the pipeline is
per-encounter, and overlapping lookup patterns are resolved by the pattern's index
in the lookup table — a property of the table, not of the data — so each block
resolves an overlap exactly as a whole-dataset run would.

Peak memory grows with `ncores`, since each worker materializes its own
long-format intermediate. On large inputs memory, not core count, is usually what
limits how high you can set it. Values above `parallel::detectCores()` are clamped
with a warning, and because forking is unavailable on Windows, `ncores > 1` there
falls back to serial with a warning.

### Choosing an AHRQ release

AHRQ revises the comorbidity software every year, and the revisions are not cosmetic: v2026.1 maps
4567 diagnosis codes where v2022.1 mapped 4319, and the mortality index weights changed between
v2022.1 and v2023.1. The same patient can therefore score differently depending on which release you
use, which matters in two situations — reproducing a published cohort, and keeping a long-running
study on one definition.

Pass `release =` to both functions to pin it:

```{r eval=FALSE}
res <- comorbidity(patient_data,
                   dx_cols  = c("dx2", "dx3", "dx4"),
                   poa_cols = c("poa2", "poa3", "poa4"),
                   release  = "2023.1")
res <- cmr_index(res, release = "2023.1")
```

The release selects three things: the diagnosis-code table, the newest ICD-10-CM version reachable
from `year`/`quarter`, and the index weights. It does **not** change the result schema — every
release produces the same 38 `CMR_*` columns, with categories a release does not define left at
zero.

`comorbidity()` records the release on its result as the `cmr_release` attribute, and `cmr_index()`
warns if you score flags from one release with another's weights:

```{r eval=FALSE}
flags <- comorbidity(patient_data, dx_cols, poa_cols, release = "2022.1")
attr(flags, "cmr_release")
#> [1] "2022.1"

cmr_index(flags)   # defaults to 2026.1 weights
#> Warning: flags were scored under AHRQ release 2022.1 but are being indexed
#> with 2026.1 weights; pass release = "2022.1" to match
```

The attribute is a warning aid, not a default — subsetting a data frame drops attributes, so
`cmr_index()` never infers the release from it.

Releases earlier than v2022.1 are not supported. AHRQ's v2021.1 is structurally different software
rather than merely older tables, and it ships no index program at all.

## Understanding Your Data

For the simple workflow above, the package used the example data you prepared.
Let's examine the structure and results in more detail.

## Detailed Analysis

```{r}
# Examine the comorbidity flags in more detail
result_summary <- result_with_indices %>%
  select(encounter_id, test_type, starts_with("CMR_")) %>%
  select(1:10)  # Show first 10 columns for brevity

print(result_summary)
```

## Risk Index Details

```{r}
# View the calculated indices
indices_summary <- result_with_indices %>%
  select(encounter_id, CMR_Index_Readmission, CMR_Index_Mortality)

print(indices_summary)

# Basic summary statistics
summary(result_with_indices[c("CMR_Index_Readmission", "CMR_Index_Mortality")])
```

## Advanced Usage

### Custom Data Sources

For advanced users who need custom lookup data, you can still provide your own formats:

```{r eval=FALSE}
# Advanced workflow with custom lookup data
# Note: These functions are internal but can still be used if needed
custom_comfmt <- build_comfmt_from_csv("path/to/your/lookup.csv")
custom_poa <- build_poa_exempt_formats("path/to/your/poa_exempt.csv")

result <- comorbidity(patient_data, 
                      dx_cols = c("dx1", "dx2", "dx3"),
                      comfmt = custom_comfmt,
                      poa_exempt = custom_poa)
```

## Understanding the Results

### Comorbidity Flags

The analysis produces 38 binary flags indicating the presence of specific comorbidities:

- **CMR_AIDS**: HIV/AIDS
- **CMR_ALCOHOL**: Alcohol abuse
- **CMR_ANEMDEF**: Deficiency anemia  
- **CMR_AUTOIMMUNE**: Autoimmune conditions
- **CMR_CANCER_METS**: Metastatic cancer
- *... and 33 others*

### Risk Indices

Two risk scores are calculated:

- **CMR_Index_Readmission**: Weighted score for readmission risk
- **CMR_Index_Mortality**: Weighted score for mortality risk

Higher scores indicate higher risk. Scores can be negative due to protective factors.

### Present on Admission (POA) Logic

The package handles POA logic automatically:

- **POA-neutral conditions**: Always assigned regardless of POA indicator
- **POA-dependent conditions**: Only assigned if POA='Y'/'W' or code is POA-exempt
- **Special cases**: CBVD logic and combination codes

## Advanced Usage

### Custom Data Sources

You can use your own comorbidity lookup files:

```{r eval=FALSE}
# Use custom lookup file
custom_comfmt <- build_comfmt_from_csv("path/to/your/lookup.csv")

# Use custom POA exempt list  
custom_poa <- build_poa_exempt_formats("path/to/your/poa_exempt.csv")
```

### Performance Considerations

For large datasets, the package automatically optimizes performance through:

- Vectorized pattern matching
- Efficient memory management
- Pre-compiled regex patterns

### Troubleshooting

Common issues and solutions:

1. **Missing POA columns**: Set `poa_cols = NULL` and `use_poa = FALSE`
2. **Wrong year/quarter**: Ensure year/quarter columns exist and contain valid values
3. **ICD code format**: The package normalizes codes automatically (removes dots, uppercase)

## References

- Healthcare Cost and Utilization Project (HCUP) Elixhauser Comorbidity Software
- Agency for Healthcare Research and Quality (AHRQ)
- CMS ICD-10-CM coding guidelines

```{r, include=FALSE}
# No cleanup needed - package handles temporary files automatically
```