---
title: "Getting Started with ecsr10"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
  pdf_document:
    toc: true
    toc_depth: 2
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Getting Started with ecsr10}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

The `ecsr10` package provides an R implementation of the Elixhauser Comorbidity Software Refined (CMR) workflow for analyzing ICD-10-CM diagnosis codes and calculating comorbidity indices. This package is designed for healthcare researchers and data analysts who need to identify comorbidities and calculate risk scores from patient diagnosis data.

It implements both families AHRQ has published, selectable per call with `variant`: the **Refined** software (`"refined"`, the default) at releases **v2021.1 through v2026.1**, and the superseded **beta** software (`"beta"`) at **v2016.2 through v2020.1**. The default release, `cmr_version()`, is Refined v2026.1. Together the two cover ICD-10-CM diagnosis codes from October 2015 through September 2026 (ICD-10-CM versions 33–43). Record the release alongside any published results: the same patient can flag differently across releases, and reproducing an existing cohort means scoring it under the release it was built with.

## Installation

Install the development version from GitHub:

```{r eval=FALSE}
# install.packages("devtools")
devtools::install_github("dpei/ecsr10")
```

```{r setup}
library(ecsr10)
library(dplyr)

cmr_releases()   # every release this package can score with
cmr_version()    # the one used when you don't say
```

## Basic Workflow

The ecsr10 package provides a simple workflow for most users:

1. **Prepare your patient diagnosis data** (wide format with diagnosis and POA columns)
2. **Apply comorbidity analysis** using the built-in lookup data
3. **Calculate risk indices** for readmission and mortality

The package automatically uses built-in lookup data, making the process straightforward.

## Step 1: Data Preparation

Your patient diagnosis data should be in wide format with:

- Diagnosis code columns (DX1, DX2, ..., DXn)
- Present on Admission (POA) columns (POA1, POA2, ..., POAn)
- Year and quarter for ICD version determination
- Patient/encounter identifiers

```{r}
# Example patient diagnosis data structure.
# dx1/poa1 are the PRINCIPAL diagnosis - present in the data, but deliberately
# not passed to comorbidity() below. See "Secondary diagnoses only".
example_patient_data <- data.frame(
  encounter_id = 1:3,
  test_type = "example",
  dx1 = c("I2510", "S72001A", "O80"),
  dx2 = c("E119",  "I10",     "Z5111"),
  dx3 = c("I10",   "E119",    "N183"),
  dx4 = c(NA,      "N183",    NA),
  poa1 = c("Y", "Y", "Y"),
  poa2 = c("Y", "Y", "Y"),
  poa3 = c("N", "Y", "W"),
  poa4 = c(NA,  "Y", NA),
  year = 2025,
  quarter = 1,
  stringsAsFactors = FALSE
)

print(example_patient_data)
```

### Secondary diagnoses only

This is the one input rule that changes your answers if you get it wrong, and it
fails quietly rather than loudly.

The AHRQ SAS program opens its diagnosis loop at position 2 —
`DO I = 2 TO MIN(&MAXNDX,&NUMDX)` — so the principal diagnosis never reaches the
comorbidity logic. `comorbidity()` scores **every column you hand it**, which
makes the exclusion your job: pass `dx2`, `dx3`, ... and leave `dx1` out. Passing
a position-1 column draws a warning, but it still scores, and the result will
differ from SAS.

Two related points:

- **`dx_cols` and `poa_cols` are paired by the first run of digits in each column
  name**, not by position in the two vectors — so `c("poa3", "poa2")` pairs
  exactly as `c("poa2", "poa3")` does. The two position sets must be identical.
  `dx3` paired with `poa4` is an error, not a silently blank POA, and so are
  duplicate positions: HCUP's own `I10_DX2` and `I10_DX3` both resolve to 10.
  Use names whose only digits are the position.
- **The encounter's diagnosis count is not applied.** SAS caps its loop at
  `I10_NDX`, ignoring a populated column beyond it. ecsr10 has no equivalent, so
  truncate `dx_cols` yourself if your data carries a meaningful count.

## Quick Start Example

```{r}
# Simple workflow - uses built-in lookup data automatically.
# Note dx1/poa1 are absent: secondary diagnoses only.
result <- comorbidity(example_patient_data,
                      dx_cols  = c("dx2", "dx3", "dx4"),
                      poa_cols = c("poa2", "poa3", "poa4"))

# Calculate risk indices
result_with_indices <- cmr_index(result)

# View summary
head(result_with_indices[,c("encounter_id", "CMR_Index_Readmission", "CMR_Index_Mortality")])
```

That's it! The package handles all the complex lookup building internally.

### Using multiple cores

`comorbidity()` takes an `ncores` argument (default `1`) that splits the
encounters into that many contiguous blocks and runs the whole pipeline on each
block in a forked worker:

```{r eval=FALSE}
result <- comorbidity(example_patient_data,
                      dx_cols  = c("dx2", "dx3", "dx4"),
                      poa_cols = c("poa2", "poa3", "poa4"),
                      ncores = 4)
```

The result is identical for every `ncores` value. Every stage of the pipeline is
per-encounter, and overlapping lookup patterns are resolved by the pattern's index
in the lookup table — a property of the table, not of the data — so each block
resolves an overlap exactly as a whole-dataset run would.

Peak memory grows with `ncores`, since each worker materializes its own
long-format intermediate. On large inputs memory, not core count, is usually what
limits how high you can set it. Values above `parallel::detectCores()` are clamped
with a warning, and because forking is unavailable on Windows, `ncores > 1` there
falls back to serial with a warning.

### Choosing an AHRQ release

AHRQ revises the comorbidity software every year, and the revisions are not cosmetic: v2026.1 maps
4567 diagnosis codes where v2022.1 mapped 4319, and the mortality index weights changed between
v2022.1 and v2023.1. The same patient can therefore score differently depending on which release you
use, which matters in two situations — reproducing a published cohort, and keeping a long-running
study on one definition.

Pass `release =` to both functions to pin it:

```{r eval=FALSE}
res <- comorbidity(patient_data,
                   dx_cols  = c("dx2", "dx3", "dx4"),
                   poa_cols = c("poa2", "poa3", "poa4"),
                   release  = "2023.1")
res <- cmr_index(res, release = "2023.1")
```

The release selects three things: the diagnosis-code table, the newest ICD-10-CM version reachable
from `year`/`quarter`, and the index weights. Every release produces 38 `CMR_*` columns in the same
order, with categories a release does not define left at zero — but **v2021.1 names two of them
differently**, `CMR_ARTH` and `CMR_CHF`, which v2022.1 renamed to `CMR_AUTOIMMUNE` and `CMR_HF`.
`comorbidity()` emits each release's own spelling, so code that hardcodes either pair has to branch
on the release.

`comorbidity()` records the release on its result as the `cmr_release` attribute, and `cmr_index()`
warns if you score flags from one release with another's weights:

```{r eval=FALSE}
flags <- comorbidity(patient_data, dx_cols, poa_cols, release = "2022.1")
attr(flags, "cmr_release")
#> [1] "2022.1"

cmr_index(flags)   # defaults to 2026.1 weights
#> Warning: flags were scored under AHRQ release 2022.1 but are being indexed
#> with 2026.1 weights; pass release = "2022.1" to match
```

The attribute is a warning aid, not a default — subsetting a data frame drops attributes, so
`cmr_index()` never infers the release from it.

### v2021.1 has no comorbidity indices

AHRQ states the Elixhauser Comorbidity Indices "are not available until v2022.1", and ships no index
program before it. `cmr_index()` therefore refuses v2021.1 rather than applying a later release's
weights:

```{r eval=FALSE}
res <- comorbidity(patient_data, dx_cols, poa_cols, release = "2021.1")
cmr_index(res, release = "2021.1")
#> Error: AHRQ ships no comorbidity index program for release "2021.1" ...
```

Score the flags under v2021.1 if that is the mapping you need, then index them explicitly under a
release that has weights — and say so in the write-up, because the two are then on different
definitions.

### The beta software (v2016.2 – v2020.1)

Before v2021.1, AHRQ published the Elixhauser Comorbidity Software for ICD-10-CM as a **beta**
version. It is superseded software rather than merely older tables:

|  | refined (v2021.1+) | beta (v2016.2 – v2020.1) |
|---|---|---|
| measures | 38 | 30, plus a derived `CMRB_HTN_C` |
| pre-existing conditions identified by | POA indicators, for 18 measures | an MS-DRG exclusion screen |
| indices | v2022.1 onward | none |
| applies to | inpatient and outpatient | inpatient only |

The MS-DRG screen suppresses a comorbidity when the encounter's MS-DRG is directly related to the
principal diagnosis — heart failure recorded on a cardiac admission is treated as the reason for the
stay rather than a pre-existing condition. That is why the beta variant takes a `drg_col`:

```{r eval=FALSE}
res <- comorbidity(patient_data,
                   dx_cols = c("dx2", "dx3", "dx4"),
                   variant = "beta",
                   release = "2020.1",
                   drg_col = "drg")
names(res)[grepl("^CMRB_", names(res))]
#> [1] "CMRB_CHF" "CMRB_VALVE" ... "CMRB_HTN_C"
```

Output columns carry a `CMRB_` prefix. That is deliberate rather than cosmetic: several beta measure
names coincide with refined ones while meaning something different — beta `CHF` is DRG-screened
where refined `CMR_HF` is POA-screened — so the two families are kept in separate namespaces, and
`cmr_index()` refuses a beta result outright.

Omitting `drg_col` runs with no screen and warns. That is faithful (AHRQ's SAS applies no
suppression when `DRG` is missing) but is rarely what an inpatient analysis wants.

Use the refined software unless you are specifically reproducing a pre-2021 analysis. AHRQ's own
advice is to use the most recent version of the tool.

## Understanding Your Data

For the simple workflow above, the package used the example data you prepared.
Let's examine the structure and results in more detail.

## Detailed Analysis

```{r}
# Examine the comorbidity flags in more detail
result_summary <- result_with_indices %>%
  select(encounter_id, test_type, starts_with("CMR_")) %>%
  select(1:10)  # Show first 10 columns for brevity

print(result_summary)
```

## Risk Index Details

```{r}
# View the calculated indices
indices_summary <- result_with_indices %>%
  select(encounter_id, CMR_Index_Readmission, CMR_Index_Mortality)

print(indices_summary)

# Basic summary statistics
summary(result_with_indices[c("CMR_Index_Readmission", "CMR_Index_Mortality")])
```

## Advanced Usage

### Custom Data Sources

Most users never need this — `release` already selects among the eleven AHRQ
tables the package ships. But `comorbidity()` accepts both lookup structures
directly, as plain R objects, if you need to score against a table of your own.

`comfmt` is a data frame of `target` and `pattern`. Under the default
`wildcard_mode = "wildcard"` a `%` in a pattern matches any run of characters;
each pattern is anchored, so `"I10"` matches only `I10`. Where two patterns match
the same code, the earlier **row** wins — order in the table is meaningful.

Targets are the AHRQ measure names **without** the `CMR_` prefix: the table
speaks `DIAB_UNCX`, the output column is `CMR_DIAB_UNCX`. A target the package
does not recognise is an error rather than a silently dropped row, so a typo
surfaces immediately.

```{r}
custom_comfmt <- data.frame(
  target  = c("DIAB_UNCX", "HTN_UNCX", "RENLFL_MOD"),
  pattern = c("E11%",      "I10",      "N183"),
  stringsAsFactors = FALSE
)

custom_result <- comorbidity(example_patient_data,
                             dx_cols  = c("dx2", "dx3", "dx4"),
                             poa_cols = c("poa2", "poa3", "poa4"),
                             comfmt   = custom_comfmt)

custom_result[, c("encounter_id", "CMR_DIAB_UNCX", "CMR_HTN_UNCX", "CMR_RENLFL_MOD")]
```

The output schema does not shrink to match: all 38 `CMR_*` columns are still
emitted, and a target your table never mentions simply stays 0.

`poa_exempt` takes the companion structure — a named list of code vectors keyed
`v33`, `v34`, ..., one entry per ICD-10-CM version, which the pipeline selects
from using each encounter's year and quarter:

```{r eval=FALSE}
custom_poa <- list(
  v42 = c("Z5111", "Z9989"),
  v43 = c("Z5111", "Z9989", "N183")
)

result <- comorbidity(example_patient_data,
                      dx_cols    = c("dx2", "dx3", "dx4"),
                      poa_cols   = c("poa2", "poa3", "poa4"),
                      comfmt     = custom_comfmt,
                      poa_exempt = custom_poa)
```

Note that `comfmt` overrides only the code table. The release's ICD-10-CM version
ceiling still applies, and the index weights still come from `release`.

## Understanding the Results

### Comorbidity Flags

The analysis produces 38 binary flags indicating the presence of specific comorbidities:

- **CMR_AIDS**: HIV/AIDS
- **CMR_ALCOHOL**: Alcohol abuse
- **CMR_ANEMDEF**: Deficiency anemia  
- **CMR_AUTOIMMUNE**: Autoimmune conditions
- **CMR_CANCER_METS**: Metastatic cancer
- *... and 33 others*

### Risk Indices

Two risk scores are calculated:

- **CMR_Index_Readmission**: Weighted score for readmission risk
- **CMR_Index_Mortality**: Weighted score for mortality risk

Higher scores indicate higher risk. Scores can be negative due to protective factors.

### Present on Admission (POA) Logic

The package handles POA logic automatically:

- **POA-neutral conditions**: Always assigned regardless of POA indicator
- **POA-dependent conditions**: Only assigned if POA='Y'/'W' or code is POA-exempt
- **Special cases**: CBVD logic and combination codes

## Performance and Troubleshooting

### Performance Considerations

For large datasets, the package automatically optimizes performance through:

- Vectorized pattern matching
- Efficient memory management
- Pre-compiled regex patterns

### Troubleshooting

Common issues and solutions:

1. **Missing POA columns**: Set `poa_cols = NULL` and `use_poa = FALSE`.

   Be aware of what this costs. 18 of the 38 measures are defined only in terms
   of POA, so they are not scoreable without it and come back as `NA` rather
   than 0 -- the same answer AHRQ's SAS program gives under `%LET POA = 0`,
   which never initialises them. The columns are still there, so the result
   binds against a POA-on one, but the liver and renal-failure hierarchies and
   the whole `CMR_CBVD` derivation do not run either. `cmr_index()` then scores
   the 20 POA-neutral measures only, and returns ordinary numbers while doing
   so -- a POA-off index looks exactly like a POA-on one and is not comparable
   with it. If you need all 38 measures, obtain POA data; the package offers
   no way to score the gated measures without it.
2. **Wrong year/quarter**: Ensure year/quarter columns exist and contain valid values
3. **ICD code format**: The package normalizes codes automatically (removes dots, uppercase)

## References

- Healthcare Cost and Utilization Project (HCUP) Elixhauser Comorbidity Software
- Agency for Healthcare Research and Quality (AHRQ)
- CMS ICD-10-CM coding guidelines

```{r, include=FALSE}
# No cleanup needed - package handles temporary files automatically
```